# Common R Commands {#sec-common-commands}

Below is a list of common commands that we use in the class, along with some examples. You can view a help file in RStudio for each command by searching for the command name in the help tab in the lower right panel. You can also just type the name of the command preceded by a “?” into the console. For example, if you wanted to understand how table works, type:

```r
?table
```

The list here does not contain information about making plots in R. That information is in @sec-plotting-cookbook. 

```{r}
#| label: load-libs-data
#| include: false
library(here)
source(here("scripts","check_packages.R"))
source(here("scripts","load_example_data.R"))
```

## Univariate statistics

### mean

Calculate the mean of a quantitative variable. Remember that this command will not work for categorical variables.

```{r}
#| label: command-mean
mean(earnings$wages)
```

None of the example datasets that we examine have missing values, but it is important to recognize that if you have missing values in a variable then the `mean` command and many of these other commands will return `NA` by default rather than a mean. To calculate the mean only for the non-missing cases, you need to add the `na.rm=TRUE` argument.

```{r}
#| label: mean-missing
mean(earnings$wages, na.rm=TRUE)
```

### median 

Calculate the median of a quantitative variable. Remember that this command will not work for categorical variables.

```{r}
#| label: command-median
median(earnings$wages)
```

### sd

Calculate the standard deviation of a quantitative variable. Remember that this command will not work for categorical variables.

```{r}
#| label: command-sd
sd(earnings$wages)
```

### IQR

Calculate the interquartile range of a quantitative variable. Remember that this command will not work for categorical variables. 

```{r}
#| label: command-iqr
IQR(earnings$wages)
```


### quantile

Calculate percentiles of a distribution. Remember that this command will not work for categorical variables. By default, the quantile command will return the quartiles (0%, 25%, 50%, 75%, and 100%). If you want different percentiles, you will have to specify the `probs` argument. 

```{r}
#| label: command-quantile
quantile(earnings$wages)
#get the 10th and 90th percentile instead
quantile(earnings$wages, probs = c(0.1,0.9))
```

### table

Calculate the absolute frequencies of the categories of a categorical variable. 

```{r}
#| label: command-table-uni
table(popularity$race)
```

### prop.table

Calculate the proportions (i.e. relative frequencies) of the categories of a categorical variable. This command must be run on the output from a `table` command. You can do that in one command by nesting the `table` command inside the `prop.table` command. 

```{r}
#| label: command-prop-table-uni
prop.table(table(popularity$race))
```

### summary

Provide a summary of a variable, either categorical or quantitative.

```{r}
#| label: command-summary-uni
summary(earnings$wages)
summary(popularity$race)
```

You can also feed in an entire dataset to a get a summary of each variable.

```{r}
#| label: command-summary-dataset
summary(earnings)
```

## Bivariate statistics

### table

the `table` command can be used to create a two-way table, although further work needs to be done to extract useful information from the two-way table. 

```{r}
#| label: command-table-bi
table(movies$genre, movies$maturity_rating)
```

### prop.table

Calculate the conditional distributions from a two-way table. The first argument here must be a two-way table output from the `table` command. It is very important that you also add a second argument that indicated the direction you want the conditional distributions. 1 will give you distributions conditional on the row variable and 2 will give you distributions conditional on the column variable. 

```{r}
prop.table(table(movies$genre, movies$maturity_rating), 1)
```

### tapply

Calculate a statistic (e.g. mean, median, sd, IQR) for a quantitative variable across the categories of a categorical variable. The first argument should be the quantitative variable. The second argument should be the categorical variable. The third argument should be the name of the command that will calculate the desired statistic.


```{r}
#| label: command-tapply
tapply(movies$runtime, movies$maturity_rating, mean)
tapply(movies$runtime, movies$maturity_rating, median)
tapply(movies$runtime, movies$maturity_rating, sd)
```


### cor

Calculate the correlation coefficient between two quantitative variables. 

```{r}
#| label: command-cor
cor(movies$rating_imdb, movies$metascore)
```


## Statistical inference

### nrow

Return the number of observations in a dataset.

```{r}
#| label: command-nrow
nrow(politics)
```

### qt 

Calculate the t-value needed for a confidence interval. For a 95% confidence interval, the first argument should always be `0.975`. The second argument should be the appropriate degrees of freedom for the statistic and dataset.

```{r}
#| label: command-qt
qt(0.975, nrow(politics)-1)
```

### pt

Calculate the p-value for a hypothesis test. The first argument should always be the negative version of the t-statistic and the second argument should be the appropriate degrees of freedom for the statistic and dataset. Remember that this will always give you the lower tail so you should multiply it by two.

```{r}
#| label: command-pt
2*pt(-2.1, nrow(politics)-1)
```


## Linear models

### lm

Run a linear model. The first argument should always be a formula of the form `dependent~independent1+independent2+...`. To simplify the writing of variable names, it is often useful to specify a second argument `data` that identifies that dataset being used. Then you don't have to include `dataset_name$` in the formula. 

Remember to always put the dependent (y) variable on the left hand side of the equation. 

```{r}
#| label: command-lm
#simple model with one independent variable
model_simple <- lm(wages~age, data=earnings)
#same simple model but recenter age on 45 years of age
model_recenter <- lm(wages~I(age-45), data=earnings)
#a model with multiple independent variables, both quantitative and qualitative
model_multiple <- lm(wages~I(age-45)+education+race+gender+nchild, data=earnings)
#a model like the previous but also with interaction between gender and nchild
model_interaction <- lm(wages~I(age-45)+education+race+gender*nchild, data=earnings)
```

Once a model object is created, information can be extracted with either the `coef` command which just reports the slopes and intercept, or a full `summary` command which gives more information.

```{r}
#| label: extract-model-estimates
coef(model_interaction)
summary(model_interaction)
```

## Miscellaneous

### round

Used for rounding the results of numbers to a given number of decimal places. By default, it will round to whole numbers, but you can specify the number of decimal places in the second argument.

```{r}
#| label: command-round
100*round(prop.table(table(movies$genre)), 3)
```

### sort

Sort a vector of numbers from smallest to largest (default), or largest to smallest (with additional argument `decreasing=TRUE`).

```{r}
#| label: command-sort
sort(100*round(prop.table(table(movies$genre)), 3), decreasing = TRUE)
sort(100*round(prop.table(table(movies$genre)), 3))
```
